---
title: "Practical Machine Learning Course Project"
output: 
  html_document:
    keep_md: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen=999)
library(lattice)
library(readr)
```
## Introduction

It is now possible to collect a large amount of data about personal
movement using activity monitoring devices such as a
[Fitbit](http://www.fitbit.com), [Nike
Fuelband](http://www.nike.com/us/en_us/c/nikeplus-fuelband), or
[Jawbone Up](https://jawbone.com/up). These type of devices are part of
the "quantified self" movement -- a group of enthusiasts who take
measurements about themselves regularly to improve their health, to
find patterns in their behavior, or because they are tech geeks. But
these data remain under-utilized both because the raw data are hard to
obtain and there is a lack of statistical methods and software for
processing and interpreting the data.

This assignment makes use of data from a personal activity monitoring
device. This device collects data at 5 minute intervals through out the
day. The data consists of two months of data from an anonymous
individual collected during the months of October and November, 2012
and include the number of steps taken in 5 minute intervals each day.

## Data

The data for this assignment can be downloaded from the course web
site:

* Dataset: [Activity monitoring data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) [52K]



The dataset is stored in a comma-separated-value (CSV) file and there
are a total of 17,568 observations in this
dataset.

<br><br>

## Loading and preprocessing the data
```{r load, echo=TRUE}
# set working directory
setwd("~/Documents/workspace/Git_Repositories/Coursera Data Science/datasciencecoursera Projects/08 Practical Machine Learning Project")

# load training dataset
pml_training <- read_csv("~/Documents/workspace/Git_Repositories/Coursera Data Science/datasciencecoursera Projects/08 Practical Machine Learning Project/Data/pml-training.csv")
```
As can be seen below, the data has been correctedly loaded into the data frame **activity**:
```{r , echo=TRUE}
str(pml_training)
```


<br><br>

